---
title: "R for Lunch"
subtitle: 'Data wrangling with {dplyr}'
author: 
  name: "John Little"
  orcid: 0000-0002-3600-0972
  affiliation:
    - name: Duke University Libraries
      department: Center for Data & Visualization Sciences
      city: Durham
      state: NC
      country: US
      url:  https://library.duke.edu/data
institute: "Center for Data & Visualization Sciences"
date: today

format: 
  revealjs:
    theme: moon
    self-contained: true  
    footer:  "[John R Little](https://JohnLittle.info) ● [Center for Data & Visualization Sciences](https://library.duke.edu/data/) ● [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/)"
    logo: images/rfun_logo.png
license: "CC BY"
editor: visual
---

## Today's topics

-   Five essential {dplyr} data wrangling verbs

-   Data pipes inside code-chunks

### Yesterday ([video](https://warpwire.duke.edu/w/bQAEAA/))

-   Import data

-   Tour of RStudio IDE

-   Coding notebooks (Quarto)

## Housekeeping

-   Drew / Lauren / breakout rooms
-   CDVS
    -   Themes
        -   Data Management [(Plans, Reproducibility, Repositories)]{style="font-size: large; color: gray;"}

        -   Data Science

        -   Data Visualization

        -   GIS and Spatial Analysis

        -   Data Sources

## Housekeeping continued

-   Website - <https://library.duke.edu/data>
-   Workshops
    -   <https://library.duke.edu/data/workshops>
-   Consulting in the Lab
    -   askData\@duke.edu
    -   my schedule: <https://is.gd/littleconsult>

## R for Lunch as a series

R for Lunch is a series that meets 8 times (till EOM Feb.) After today it will meet regularly on Thursdays at noon.

-   Sign-up for each workshop individually

-   Each episode has a unique zoom link

## Eat yor own dog food

\
Model how R can work for practical reproducible workflows

-   Code in RStudio

-   One kind of report is these slides ([GitHub](https://github.com/libjohn/rforlunch_exercises/blob/main/slides/slides_import-data.pdf "PDF report - slides"))

-   Another report is the [*Introduction to R/Tidyverse/Quarto* text](https://intro2r.library.duke.edu/).

# Definitions

### Opinionated

> Tidyverse and Quarto is the most practical and well developed, reproducible, scientific analysis and publishing workflow available.

### Tidyverse and Tidy data is an important foundation.

-   Every row is a single observation
-   Every column is a variable
-   The cells are single data values

# Tidy data

## Tidy data[^1] {background-image="images/tidy_data.png" background-size="contain" background-repeat="false"}

[^1]: A robust discussion of *tidy data* can be found in *R for Data Science* ([Wickham, Çetinkaya-Rundel, and Grolemund 2023](https://intro2r.library.duke.edu/longer_wider#ref-wickham2023))*: [https://r4ds.had.co.nz/tidy-data.html](https://r4ds.had.co.nz/tidy-data.htmlhttps://r4ds.had.co.nz/tidy-data.html)*

## Tidy data

-   Every row is a single observation
-   Every column is a variable
-   The cells are single data values

## Wide data

```{r}
#| message: false
#| warning: false
#| code-line-numbers: "4"
#| echo: true
#| code-fold: true
library(tidyverse)
library(gt)
library(gtExtras)

tidyr::relig_income |> 
  gt::gt_preview() |> 
  gtExtras::gt_theme_dark()
```

## Tall data

::: columns
::: {.column width="40%"}
```{r}
#| code-line-numbers: 2-4
#| echo: true
#| code-fold: true
relig_income |> 
  pivot_longer(cols = -religion, 
               names_to = "income_category", 
               values_to = "income") |> 
  gt::gt_preview() |> 
  gtExtras::gt_theme_dark()
```
:::

::: {.column width="60%"}
```{r}
#| code-line-numbers: "2,7-9,11"
#| echo: true
#| code-fold: true
#| fig-height: 8
relig_income |> 
  pivot_longer(cols = -religion, 
               names_to = "income_category", 
               values_to = "income") |> 
  mutate(religion = fct_relevel(religion, "Evangelical Prot", "Mainline Prot", "Catholic", "Unaffiliated", "Historically Black Prot")) |> 
  mutate(income_category = fct_rev(as_factor(income_category))) |>
  ggplot(aes(income, income_category)) +
  geom_col(fill = "#eee8d5") +
  facet_wrap(vars(
    fct_other(
      religion, 
      keep = c("Evangelical Prot", "Mainline Prot", "Catholic", "Unaffiliated", "Historically Black Prot")))) +
  theme(plot.background = element_rect(fill = "#002b36"),
        text = element_text(color = "#eee8d5"),
        axis.text = element_text(color = "#eee8d5"), 
        panel.background = element_rect(fill = "#002b36"),
        panel.grid = element_line(color = "#002b36"),
        strip.background = element_rect(fill = "#7b9c9f"))
                                 
```
:::
:::

## Code

 \

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "2,5|3-4"
relig_income |> 
  pivot_longer(cols = -religion, names_to = "income_category") |> 
  ggplot(aes(value, income_category)) +
  geom_col() +
  facet_wrap(vars(religion))
```

::: {style="color: dimgray; font-size: 35%; text-align: center;"}
Image Credit: apreshill \| CC BY 4.0 \| https://github.com/apreshill/teachthat/blob/master/pivot/pivot_longer_smaller.gif\]
:::

# Polls

# We are here to help

-   askData\@duke.edu

-   https://library.duke.edu/data

-   https://is.gd/littleconsult

# Let's do it

## Two things for today

-   five essential {dplyr} data wrangling verbs

-   data pipes inside code-chunks

-   <https://intro2r.library.duke.edu/wrangle.html>

## Exercises

1.  https://intro2r.library.duke.edu/ \> Exercises \> Link out \> Green [**Code**]{style="color:green"} button \> Download ZIP

2.  Then, Unzip (i.e. Expand) the folder (on your local file system)

3.  Then, double click the `rforlunch_exercises.Rproj` file

4.  From **RStudio** the Files tab, open the `01_dplyr.qmd`

    -   The answer file is in the **RStudio** `rforlunch_exercises` project \> `Files Tab > Answers folder`

# Closing

## Pipes and Assignments

 

+--------------+---------------+--------------+-------------------+
| Operator     | Operator Name | Keystore     | Pnuemonic         |
+==============+===============+==============+===================+
| `<-`         | assignment    | Alt-dash     | "Gets value from" |
+--------------+---------------+--------------+-------------------+
| `|>`\        | pipe          | Ctrl-Shift-M | "And then"        |
| or           |               |              |                   |
|              |               |              |                   |
| `%>%`        |               |              |                   |
+--------------+---------------+--------------+-------------------+

## Citation management

 

RStudio \> Quarto Notebook \> Insert \> Citation

Example DOI: 10.18637/jss.v059.i10

## ai-paired coding

 

-   Data science concepts: **Microsoft copilot** ("More precise" setting)

-   Code completion: **GitHub copilot** and RStudio (IDE) or VSCode (IDE)

# Bye for now

-   askData\@duke.edu
-   <https://is.gd/littleconsult>
-   <https://library.duke.edu/data>
